---
title: "chapter4"
author: "Laura Jernstöm"
date: "11/18/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Chapter 4

```{r message=FALSE}

library(MASS)
library(tidyr)

data(Boston)

str(Boston)
dim(Boston)
summary(Boston)


```
The Boston data set contains information on housing Values in suburbs of Boston and includes 506 observations of 14 variables. 
There are information including crime rate, proportion on residential land zoned for lots, nitrogen oxiden concentration, age and size of living units, residential taxes and pupil-teacher ratio.

```{r message=FALSE}

#install.packages("corrplot")
library(corrplot)

pairs(Boston)

cor_matrix<-cor(Boston) 

# print the correlation matrix
cor_matrix %>% round(digits=2)

# visualize the correlation matrix
corrplot(cor_matrix, method="circle", type = "upper", cl.pos = "b", tl.pos = "d", tl.cex = 0.6)



```

Describe...


```{r message = TRUE}

# 4.

boston_scaled <- scale(Boston)

# summaries of the scaled variables
summary(boston_scaled)

# class of the boston_scaled object
class(boston_scaled)

# change the object to data frame
boston_scaled <- as.data.frame(boston_scaled)


```

The change to previous can be seen from looking at the summary statistics of the variables: The means are now standardised to to zero.

```{r}

# 4. continues

# Categorical variable crime rate and division into training and testing sets.

summary(boston_scaled$crim)

# create a quantile vector of crim and print it


bins <- quantile(boston_scaled$crim)
bins

# create a categorical variable 'crime'
crime <- cut(boston_scaled$crim, breaks = bins, include.lowest = TRUE, labels = c("low", "med_low", "med_high", "high"))

# look at the table of the new factor crime
table(crime)

# remove original crim from the dataset
boston_scaled <- dplyr::select(boston_scaled, -crim)

# add the new categorical value to scaled data
boston_scaled <- data.frame(boston_scaled, crime)


## Division into training and testing

# number of rows in the Boston dataset 
n <- nrow(boston_scaled)

# choose randomly 80% of the rows
ind <- sample(n,  size = n * 0.8)

# create train set
train <- boston_scaled[ind,]

# create test set 
test <- boston_scaled[-ind,]

# save the correct classes from test data
correct_classes <- test$crime

# remove the crime variable from test data
test <- dplyr::select(test, -crime)




```

Comments....


```{r message = FALSE}

# 5. 

# MASS and train are available

# linear discriminant analysis
lda.fit <- lda(crime ~ ., data = train)

# print the lda.fit object
lda.fit

# the function for lda biplot arrows
lda.arrows <- function(x, myscale = 1, arrow_heads = 0.1, color = "orange", tex = 0.75, choices = c(1,2)){
  heads <- coef(x)
  arrows(x0 = 0, y0 = 0, 
         x1 = myscale * heads[,choices[1]], 
         y1 = myscale * heads[,choices[2]], col=color, length = arrow_heads)
  text(myscale * heads[,choices], labels = row.names(heads), 
       cex = tex, col=color, pos=3)
}

# target classes as numeric
classes <- as.numeric(train$crime)

# plot the lda results
plot(lda.fit, dimen = 2, col = classes, pch = classes)
lda.arrows(lda.fit, myscale = 1)


```


```{r message = FALSE}

# 6. 

# predict classes with test data
lda.pred <- predict(lda.fit, newdata = test)

# cross tabulate the results
table(correct = correct_classes, predicted = lda.pred$class)



```

Comment....


```{r message = FALSE}

# 7. Distance measures (K-mass, euclidian, manhattan)


library(MASS)
data('Boston')

# euclidean distance matrix
dist_eu <- dist(Boston)

# look at the summary of the distances
summary(dist_eu)

# manhattan distance matrix
dist_man <- dist(Boston, method = "manhattan")

# look at the summary of the distances
summary(dist_man)

# Scale the variables
boston_scaled <- scale(Boston)

# summaries of the scaled variables
summary(boston_scaled)

# class of the boston_scaled object
class(boston_scaled)

# change the object to data frame
boston_scaled <- as.data.frame(boston_scaled)

# k-means clustering
km <- kmeans(boston_scaled, centers = 3)

# plot the Boston dataset with clusters
p <- pairs(boston_scaled[6:10], col = km$cluster)
p


```

Comment...



```{r message = FALSE}

# 7. Continues

# MASS, ggplot2 and Boston dataset are available
set.seed(123)

# determine the number of clusters
k_max <- 10

# calculate the total within sum of squares
twcss <- sapply(1:k_max, function(k){kmeans(boston_scaled, k)$tot.withinss})

# visualize the results
qplot(x = 1:k_max, y = twcss, geom = 'line')

# k-means clustering
km <-kmeans(boston_scaled, centers = 2)

# plot the Boston dataset with clusters
pairs(boston_scaled, col = km$cluster)




```

Interpret...



