---
title: "chapter6"
author: "Laura Jernstöm"
date: "12/2/2020"
output: html_document
---


# CHAPTER 6 


## Part 1.

Analysis of Chpater 8 of Kimmos' book with rats data

Description of data:

"Data from a nutrition study conducted in three groups of rats
(Crowder and Hand, 1990). The three groups were put on different diets, and
each animal’s body weight (grams) was recorded repeatedly (approximately
weekly, except in week seven when two recordings were taken) over a 9-week
period." (Vehkalahti and Everitt, ?)


Implement the analyses of Chapter 8 of MABS using the RATS data. (0-7 points: 0-4 points for graphs or analysis results + 0-3 points for their interpretations)



```{r}

rm(list=ls())

date()



# Data
rats <- read.table("https://raw.githubusercontent.com/KimmoVehkalahti/MABS/master/Examples/data/rats.txt",
                   sep ="", header = TRUE)
ratsl <- read.table( "Z:/FDPE/Courses/Datamooc/IODS-project/data/ratsl.txt", sep = " ", header = TRUE)

names(ratsl)


#Access the package ggplot2
library(ggplot2)
library(tidyr)
library(dplyr)

ratsl$ID <- as.factor(ratsl$ID)
ratsl$Group <- as.factor(ratsl$Group)

# Draw the plot
ggplot(ratsl, aes(x = Time, y = rats, linetype = ID)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ Group, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(limits = c(min(ratsl$rats), max(ratsl$rats))) + ggtitle("Figure 1")




```

Figure 1 plots the individual response profiles of by treatment group for the rats data. First thing
to notice are the clear level differences between the groups: group 2 is slightly below in level terms
to group 3 at all values of Time-variable, where as group 1 is significantly below groups 2 and 3 with
a difference 300 points from the max values of group 1 vs minimum of group 2. 

All groups seam to have relatively similar slopes among individuals, with perhaps a small exception in Group 2,
which speaks of tracking phenomenon i.e. individuals having high scores in the beginning tend to have them
in the end as well.

Variability in the groups looks quite small with all groups having one possible outlier with Group 2 having one
large outlier. Slopes are also increasing i.e. rats scores increase over time.


```{r}

# Standardise the variable rats
ratsl <- ratsl %>%
  group_by(Time) %>%
  mutate(stdrats = (rats - mean(rats))/sd(rats) ) %>%
  ungroup()

# Glimpse the data
glimpse(ratsl)

# Plot again with the standardised bprs
ggplot(ratsl, aes(x = Time, y = stdrats, linetype = ID)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ Group, labeller = label_both) +
  scale_y_continuous(name = "standardized rats") + ggtitle("Figure 2")


```

After having standardized the variable rats, the figures look quite different. Comparing Figure 1 and Figure 2,
first thing to notice is the change in the slopes to the opposite, which seams a bit strange. It still looks
like all groups have possible outlier with Group 2 having a clear outlier.



```{r}

# Number of time periods, baseline (time 0) included
n <- ratsl$Time %>% unique() %>% length()


# Summary data with mean and standard error of rats by Group and Time 
ratsls <- ratsl %>%
  group_by(Group, Time) %>%
  summarise( mean = mean(rats), se = sd(rats)/sqrt(n) ) %>%
  ungroup()

# Glimpse the data
glimpse(ratsls)

# Plot the mean profiles
ggplot(ratsls, aes(x = Time, y = mean, linetype = Group, shape = Group)) +
  geom_line() +
  scale_linetype_manual(values = c(1,2,3)) +
  geom_point(size=3) +
  scale_shape_manual(values = c(1,2,3)) +
  geom_errorbar(aes(ymin = mean - se, ymax = mean + se, linetype="1"), width=0.3) +
  theme(legend.position = c(0.9,0.5)) +
  scale_y_continuous(name = "mean(rats) +/- se(rats)") + ggtitle("Figure 3")


```

Figure 3 plots the mean weights of the rats against time by group along with their confidence intervals.
There is no overlapping i.e all groups are significantly different to each other.






```{r}

# Create a summary data by treatment and subject with mean as the summary variable (ignoring baseline week 0).
rats8l <- ratsl %>%
  filter(Time > 0) %>%
  group_by(Group, ID) %>%
  summarise( mean=mean(rats) ) %>%
  ungroup()

# Glimpse the data
glimpse(rats8l)

# Draw a boxplot of the mean versus treatment
ggplot(rats8l, aes(x = Group, y = mean)) +
  geom_boxplot() +
  stat_summary(fun.y = "mean", geom = "point", shape=23, size=4, fill = "white") +
  scale_y_continuous(name = "mean(rats), Time") + ggtitle("Figure 4")

# Create a new data by filtering the outlier and adjust the ggplot code the draw the plot again with the new data
rats8l1 <- rats8l %>%
  filter(mean < 60)




```

Figure 4 shows a clear difference in means between the three groups i.e. the development of the
weights of the rats are significantly different for the different diets. Figure 4 also supports the notion of one outlier for each group, with the outlier in Group 2 the most clear.


```{r}

# Perform a three-sample f-test?
# t.test(mean ~ Group, data = rats8l1, var.equal = TRUE)

# Add the baseline from the original data as a new variable to the summary data
rats8l2 <- rats8l %>%
  mutate(baseline = rats$WD1)

# Fit the linear model with the mean as the response 
fit <- lm(mean ~ baseline + Group, data = rats8l2)


# Compute the analysis of variance table for the fitted model with anova()
anova(fit)


?f.test

```

Looking at the table showing the results from the analysis of variance, we see that the both the baseline
and group have significant estimates indicating clear differences from the null of zero difference in means.




## Part 2.


rm(list=ls())

date()



# Data
rats <- read.table("https://raw.githubusercontent.com/KimmoVehkalahti/MABS/master/Examples/data/rats.txt",
                   sep ="", header = TRUE)
ratsl <- read.table( "Z:/FDPE/Courses/Datamooc/IODS-project/data/ratsl.txt", sep = " ", header = TRUE)

names(ratsl)


#Access the package ggplot2
library(ggplot2)
library(tidyr)
library(dplyr)

ratsl$ID <- as.factor(ratsl$ID)
ratsl$Group <- as.factor(ratsl$Group)




